### 2025-07-14 manuscript/part1.tex
introduction and parser

\chapter{Introduction}
\section{Motivation and Problem Statement}

CPS team - waht do they need it for
code growing user base - interconnecticty os , other qualities speaking for the usage of vs code
-
already in possesion of Haskell Checker but

\label{sec:scope}
\section{Objectives and Scope}
The main goal of the thesis was to create a \textit{Language Extension} ?NAME? for \textit{MiniProb}. Language extensions are a big part of
VS Code’s extension ecosystem and enable the editor to utilize custom language tooling. Such extensions support the user during the implementation process
while writing code, by providing language assistance like syntax validation or code completion. While this implementation targets VS Code, the underlying
logic and functionality are not limited to it - through the use of the \textit{Language Server Protocol (LSP)}, the developed tooling can be reused across
various editors and IDEs that support the protocol. The core functionality of these extensions stems from \textit{parsers} which, against a formal language
definition, convert written text into abstracted parts, validating the code in the process. These parts can then be used to extend the capabilities of the
tooling to encompass referential validation, type checking, code completion, diagnostics, and other language services that enhance the development experience.

In order for parsers to function accordingly, they require formal language definitions or models. These definitions range from various types of grammars to
abstract machines such as automata, which together provide the structural and syntactic rules necessary for correct interpretation and validation of source code.
Based on these definitions, parsers are able to systematically identify the hierarchical structure of a program by recognizing patterns, token sequences, and nested
constructs, ensuring that the code adheres to the expected form before any further processing or analysis occurs.
In this thesis, a generative context-free grammar is used to formally describe \textit{MiniProb}, a domain-specific language (DSL) designed for authoring \textit{POMC} files.
With this grammar as a foundation, parsers enable the implementation of a fully functional language extension to aid developers writing POMC files.\\
Ultimately, the resulting extension, titled ?NAME?, is intended to provide comprehensive language support. This includes syntactic and semantic analysis
through syntax highlighting and validation, accurate resolution of symbol references and rudimentary code completion, as well as implemented type checking mechanisms
to ensure compatibility during development.

@@

Lastly, this thesis includes an evaluation of the newly implemented parser, focusing on its correctness and performance in practical usage scenarios.
The evaluation is based on metrics collected from representative input samples, measuring factors such as parsing speed, memory usage, and error handling.
Where relevant, comparisons are also drawn against the existing \textit{Haskell}-based parser for \textit{.pomc} files, offering a point of reference to assess
improvements or trade-offs introduced by the new implementation.

@@

\subsection{Parsing Models}

>>>>

\chapter{Introduction}
\section{Motivation}

Created by the <Faculty ()CPS> - .... formal verification, model checking, temporal logics and automata - the MiniProb DSL provides a concise notation
for probabilistic process models\cite{guideMiniProb}. At present, MiniProb tooling is limited to a Haskell-based parser, yielding a tedious workflow: source code is written
in a generic editor and syntax errors discovered one at a time as the parser aborts on its first failure. A modern editor extension - for example, a
Visual Studio Code plugin providing syntax highlighting and live diagnostics - would dramatically streamline development. Such an interface lowers the barrier to entry for
new users and also offers a familiar environment in which MiniProb programs can be authored and the further automatically converted into automata for formal verification \cite{2025modelcheckingprobabilisticoperator,POPACheck}.
In the context of cyber-physical systems - where interconnected devices underpin critical infrastructure - and the ever-growing complexity and threat landscape of cyberspace,
integrated verification support is essential to ensure system reliability, preempt security vulnerabilities, and uphold correctness guarantees under rigorous temporal and
probabilistic constraints.

\label{sec:scope}
\section{Objectives and Scope}
The main goal of the thesis was to create a \textit{Language Extension} ?NAME? for \textit{MiniProb}. Language extensions are a big part of
VS Code's extension ecosystem and enable the editor to utilize custom language tooling. Such extensions support the user during the implementation process
while writing code, by providing language assistance like syntax validation or code completion. While this implementation targets VS Code, the underlying
logic and functionality are not limited to it - through the use of the \textit{Language Server Protocol (LSP)}, the developed tooling can be reused across
various editors and IDEs that support the protocol. The core functionality of these extensions stems from \textit{parsers} which, against a formal language
definition, convert written text into abstracted parts \ref{sec:parser}.

In this thesis, a generative context-free grammar is used to formally describe \textit{MiniProb} \ref{sec:miniprob}, a domain-specific language (DSL) designed for authoring \textit{POMC} files.
With this grammar as a foundation, parsers enable the implementation of a fully functional language extension to aid developers writing POMC files.\\
Ultimately, the resulting extension, titled ?NAME?, is intended to provide comprehensive language support. This includes syntactic and semantic analysis
through syntax highlighting and validation, accurate resolution of symbol references and rudimentary code completion, as well as implemented type checking mechanisms
to ensure compatibility during development.

@@

Finally, the complete source code is hosted on GitHub at \url{https://github.com/e12224207/miniprob}, and prebuilt \code{.vsix} packages can be downloaded from the Releases page for straightforward installation.
This documentation serves as the definitive guide to the application developed in that repository, underpinned by the theoretical foundations of service customization and language extension principles.

@@

\subsection{Parsing Models}
\label{sec:parser}

The core functionality of language extensions stems from \textit{parsers} which, against a formal language
definition, convert written text into abstracted parts, validating the code in the process. These parts can then be used to extend the capabilities of the
tooling to encompass referential validation, type checking, code completion, diagnostics, and other language services that enhance the development experience.
In order for parsers to function accordingly, they require formal language definitions or models. These definitions range from various types of grammars to
automata such as PDAs, which provide the structural and syntactic rules necessary for correct interpretation and validation of source code.
Based on these definitions, parsers are able to systematically identify the hierarchical structure of a program by recognizing patterns, token sequences, and nested
constructs, ensuring that the code adheres to the expected form before any processing or analysis of the code occurs.
With a defined grammar as a foundation, parsers enable the implementation of fully functional development tools - for example, a language extension that assists with code editing.
\\
